# Summary {#LogRegSummary}



The second data frame represents the presence or absence of a common shrub (*Berberis repens*) in the [Bryce Canyon National Park](https://www.nps.gov/brca/index.htm) Utah. The researchers recorded whether the shrub was present or not in many 10m&times;10m plots throughout the park. They also recorded the elevation (among other things) of the plot. They wanted to model the effect of elevation on the presence of the shrub. Their data is in [BryceShrub.csv](https://derekogle.com/Book207/data/BryceShrub.csv), which are loaded below.

```{r}
shrub <- read.csv("https://derekogle.com/Book207/data/BryceShrub.csv")
```

For both data sets an indicator variable for the response variable is needed to make plotting the logistic regression results easier. This variable is created with `ifelse()` which has a "condition" as its first argument, what should be returned if the condition is "yes" as the second argument, and what should be returned if the condition is "no" as the third argument. To be consistent with R the condition should always be asking for the "second" level of the variable, which will be the alphabetically second level if the levels have not be altered by you.

Thus, for the `putts` data there are two levels of the `made` variable so that a new variable called `made01` is "1" if `made` is "Yes" and "0" if it is a "No."

```{r}
unique(putts$made)  # Available levels
putts$made01 <- ifelse(putts$made=="Yes",1,0)
headtail(putts)
```

A similar process if followed to create a `presence01` variable in the `shrub` data frame.

```{r}
unique(shrub$presence)  # Available levels
shrub$presence01 <- ifelse(shrub$presence=="Present",1,0)
headtail(shrub)
```
