# Appendix -- R Generic Code{-}

## Common Codes{-}
The generic codes below are provided here so that you can efficiently copy and paste them into your assignment. They are generally ordered by module withing major conceptual topics. Most code that is in ALL CAPITALS should be replaced with something more specific (with the exception of `TRUE` and `FALSE`). The following are fairly common genertic codes that should be replaced. Other items are described below. All other code should be typed (preferably copied-and-pasted) as shown, though you may change the name of the assigned object to the left of the `<-`.

* `DFOBJ` should be replaced with the name of your data frame.
* `QVAR` should be replaced with the name of your quantitative variable.
    * `QVARR` for response and `QVARE` for explanatory variables.
* `CVAR` should be replaced with the name of your categorical variable.
* `NUM` should be replaced with a number.
* `XXINDIVIDUALXX` should be replaced with what an individual is.
* `XXQVAR LABELXX` should be replaced with a descriptive label for `QVAR`.
* `XXCVAR LABELXX` should be replaced with a descriptive label for `CVAR`.


## Foundations{-}
#### 2-Sample t-test{-}

* `HA` is replaced with `"two.sided"` for not equals, `"less"` for less than, or `"greater"` for greater than alternative hypotheses (H<sub>A</sub>).
* `CNFVAL` is the confidence level as a proportion (e.g., `0.95`).

```{r eval=FALSE, highlight=FALSE}
## Levene's Test
levenesTest(QVAR~CVAR,data=DFOBJ)

## 2-Sample t-test
t.test(QVAR~CVAR,data=DFOBJ,alt=HA,conf.level=CNFVAL,var.equal=TRUE)

## Histogram Separated Into Groups
ggplot(data=DFOBJ,mapping=aes(x=QVAR)) +
  geom_histogram(binwidth=NUM,boundary=0,color="black",fill="lightgray") +
  labs(x="XXQVAR LABELXX",y="Frequency of XXINDIVIDUALSXX") +
  scale_y_continuous(expand=expansion(mult=c(0,0.05))) +
  theme_NCStats() +
  facet_wrap(vars(CVAR))

## Construct Summary Graphic of Points, Mean, and CIs
ggplot(data=DFOBJ,mapping=aes(x=CVAR,y=QVAR)) +
  geom_jitter(alpha=0.5,width=0.05) +
  stat_summary(fun.data=mean_cl_normal,geom="errorbar",size=2,width=0) +
  stat_summary(fun=mean,geom="point",pch=21,fill="white",size=2) +
  labs(x="XXCVAR LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()
```

#### Model Comparisons{-}

```{r eval=FALSE, highlight=FALSE}
## Fit the Linear Model
aov1 <- lm(QVAR~CVAR,data=DFOBJ)

## Extract the ANOVA Table
anova(aov1)

## Extract Table of Model Coefficients
cbind(ests=coef(aov1),confint(aov1))
```

#### Compute p-value from F test statistic{-}

* `FVAL` should be replaced with the calculated F test statistic.

```{r eval=FALSE, highlight=FALSE}
distrib(FVAL,distrib="f",df1=NUM,df2=NUM,lower.tail=FALSE)
```

&nbsp;

## One-Way ANOVA{-}
#### Basic One-Way ANOVA{-}

```{r eval=FALSE, highlight=FALSE}
## Fit the Linear Model
aov1 <- lm(QVAR~CVAR,data=DFOBJ)

## Extract the ANOVA Table
anova(aov1)

## Construct Summary Graphic
## !!!! Only use if no multiple comparisons or transformation !!!!
ggplot(data=DFOBJ,mapping=aes(x=CVAR,y=QVAR)) +
  geom_jitter(alpha=0.5,width=0.05) +
  stat_summary(fun.data=mean_cl_normal,geom="errorbar",size=2,width=0) +
  stat_summary(fun=mean,geom="point",pch=21,fill="white",size=2) +
  labs(x="XXCVAR LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()
```

#### Multiple Comparisons{-}
This assumes that you fit a one-way ANOVA linear model using `lm()` and assigned the results to `aov1`.

```{r eval=FALSE, highlight=FALSE}
## Tukey's Results
mc <- emmeans(aov1,specs=pairwise~CVAR)
( mcsum <- summary(mc,infer=TRUE) )

## Construct Summary Graphic
## !!!! Only use if no transformation !!!!
ggplot() +
  geom_jitter(data=DFOBJ,mapping=aes(x=CVAR,y=QVAR),
              alpha=0.5,width=0.05) +
  geom_errorbar(data=mcsum$emmeans,
                mapping=aes(x=CVAR,ymin=lower.CL,ymax=upper.CL),
                size=2,width=0) +
  geom_point(data=mcsum$emmeans,mapping=aes(x=CVAR,y=emmean),
             size=2,pch=21,fill="white") +
  labs(x="XXCVAR LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()
```

#### Assumption Checking{-}
This assumes that you fit a one-way ANOVA linear model using `lm()` and assigned the results to `aov1`.

```{r eval=FALSE, highlight=FALSE}
## Create asummption checking graphic
assumptionCheck(aov1)

## Show croup sample sizes
xtabs(~CVAR,data=DFOBJ)
```

#### Transformations{-}

```{r eval=FALSE, highlight=FALSE}
## Create assummption checking graphic with lambda transformation
## !!!! lambday= 0, 1/2, 1/3, 1/4, -1/2, -1/3, or -1/4
assumptionCheck(aov1,lambday=NUM)
```

The following are codes for possible transformations of the response variable. Use **one** of these after making a decision from the previous code.

```{r eval=FALSE, highlight=FALSE}
DFOBJ$tQVAR <- log(DFOBJ$QVAR)      # natural log transformation
DFOBJ$tQVAR <- sqrt(DFOBJ$QVAR)     # square root transformation
DFOBJ$tQVAR <- DFOBJ$QVAR^(1/3)     # cube root transformation
DFOBJ$tQVAR <- DFOBJ$QVAR^(1/4)     # fourth root transformation
DFOBJ$tQVAR <- 1/DFOBJ$QVAR         # inverse (or reciprocal) transformation
DFOBJ$tQVAR <- DFOBJ$QVAR^(-1/2)    # inverse square root transformation
DFOBJ$tQVAR <- DFOBJ$QVAR^(-1/3)    # inverse cube root transformation
DFOBJ$tQVAR <- DFOBJ$QVAR^(-1/4)    # inverse fourth root transformation
```

```{r eval=FALSE, highlight=FALSE}
## Fit new linear model with transformed response variable (note tQVAR)
aov2 <- lm(tQVAR~CVAR,data=DFOBJ)
anova(aov2)
```

Multiple comparisons on the TRANSFORMED scale are made below. See code further below for multiple comparisons on the back-transformed scale (when using a log or square root transformation).

```{rr eval=FALSE, highlight=FALSE}
## on transformed scale (must use this for non log or square root)
mct <- emmeans(aov2,specs=pairwise~CVAR)
( mcsumt <- summary(mct,infer=TRUE) )               # transformed scale

## summary graphic on transformed scale
ggplot() +
  geom_jitter(data=DFOBJ,mapping=aes(x=CVAR,y=tQVAR),
              alpha=0.5,width=0.05) +
  geom_errorbar(data=mcsumt$emmeans,
                mapping=aes(x=CVAR,ymin=lower.CL,ymax=upper.CL),
                size=2,width=0) +
  geom_point(data=mcsumt$emmeans,mapping=aes(x=CVAR,y=emmean),
             size=2,pch=21,fill="white") +
  labs(x="XXCVAR LABELXX",y="XXtQVAR LABELXX") +
  theme_NCStats()
```

Multiple comparisons on the BACK-TRANSFORMED scale are made below (only for log and square root transformations).

```{r eval=FALSE, highlight=FALSE}
## only after log transformation
mct <- emmeans(aov2,specs=pairwise~CVAR,tran="log")
( mcsumbt <- summary(mct,infer=TRUE,type="response") )    # back-transformed scale

## only after square root transformation
mct <- emmeans(aov2,specs=pairwise~CVAR,tran="sqrt")
( mcsumbt <- summary(mct,infer=TRUE,type="response") )    # back-transformed scale

## summary graphic on back-transformed scale
ggplot() +
  geom_jitter(data=DFOBJ,mapping=aes(x=CVAR,y=QVAR),
              alpha=0.5,width=0.05) +
  geom_errorbar(data=mcsumbt$emmeans,
                mapping=aes(x=CVAR,ymin=lower.CL,ymax=upper.CL),
                size=2,width=0) +
  geom_point(data=mcsumbt$emmeans,mapping=aes(x=CVAR,y=response),
             size=2,pch=21,fill="white") +
  labs(x="XXCVAR LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()
```

&nbsp;

## Two-Way ANOVA{-}
#### Basic Two-Way ANOVA{-}

```{r eval=FALSE, highlight=FALSE}
## Fit the Linear Model
aov2 <- lm(QVAR~CVAR1+CVAR2+CVAR1:CVAR2,data=DFOBJ)

## Assumption checking
xtabs(~CVAR1+CVAR2,data=DFOBJ)
assumptionCheck(aov2)

## Extract the ANOVA Table
anova(aov2)
```

#### Multiple Comparisons{-}
If there IS an interaction effect then computed Tukey's results for the interaction term in `aov2` (see below for when there is no interaction effect).

```{r eval=FALSE, highlight=FALSE}
## !!!! Only if there is a significant interaction effect !!!!
## !!!!!! Only if no transformation !!!!
mc2 <- emmeans(aov2,specs=pairwise~CVAR1:CVAR2)
( mc2sum <- summary(mc2,infer=TRUE) )

## !!!!!! Only if log transformation !!!!
mc2t <- emmeans(aov2,specs=pairwise~CVAR1:CVAR2,tran="log")
( mc2sumt <- summary(mc2t,infer=TRUE) )  # still transformed
( mc2sumbt <- summary(mc2t,infer=TRUE,type="response") )  # back-transformed
```

If there is NO interaction effect then fit a new model without the interaction term.

```{r eval=FALSE, highlight=FALSE}
## !!!! Only if there is no significant interaction effect !!!!
aov2_noint <- lm(QVAR~CVAR1+CVAR2,data=DFOBJ)
```

Then compute Tukey's results for the main effect terms that are significant. Note you may need to repeat the code below, once for `CVAR1` and then again for `CVAR2`.

```{r eval=FALSE, highlight=FALSE}
## !!!! Only if there is no significant interaction effect !!!!
## !!!!!! Only if no transformation !!!!
mc2_noint <- emmeans(aov2_noint,specs=pairwise~CVAR1)
( mc2sum_noint <- summary(mc2_noint,infer=TRUE) )

## !!!!!! Only if log transformation !!!!
mc2t_noint <- emmeans(aov2_noint,specs=pairwise~CVAR1,tran="log")
( mc2sumt_noint <- summary(mc2t_noint,infer=TRUE) )  # still transformed
( mc2sumbt_noint <- summary(mc2t_noint,infer=TRUE,type="response") )  # back-trans
```

#### Interaction and Main Effects Plots{-}

```{r eval=FALSE, highlight=FALSE}
## Interaction plot ... may switch CVAR1 and CVAR2
pd <- position_dodge(width=0.1)  ## may try different values
ggplot(data=mc2sum$emmeans,mapping=aes(x=CVAR1,group=CVAR2,color=CVAR2,
                                       y=emmean,ymin=lower.CL,ymax=upper.CL)) +
  geom_line(position=pd,size=1.1,alpha=0.25) +
  geom_errorbar(position=pd,size=2,width=0) +
  geom_point(position=pd,size=2,pch=21,fill="white") +
  labs(x="XXCVAR1 LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()

## Main effects plot ... may need CVAR2
## !!!! Only use if there is no significant interaction effect !!!!
ggplot(data=mc2sum_noint$emmeans,
       mapping=aes(x=CVAR1,group=1,y=emmean,ymin=lower.CL,ymax=upper.CL)) +
  geom_line(size=1.1,alpha=0.25) +
  geom_errorbar(size=2,width=0) +
  geom_point(size=2,pch=21,fill="white") +
  labs(x="XXCVAR1 LABELXX",y="XXQVAR LABELXX") +
  theme_NCStats()
```

Note that you may need to use transformations and, thus, you may need to use code similar to what was shown above for the One-Way ANOVA (remember to replace `y=emmean` with `y=response` when using back-transformed results).

&nbsp;

## Simple Linear Regression{-}
#### Basic Linear Regression{-}
```{r eval=FALSE, highlight=FALSE}
## Fit the linear model
( slr <- lm(QVARR~QVARE,data=DFOBJ) )

## Make a basic prediction (without confidence/prediction intervals)
predict(slr,newdata=data.frame(QVARE=NUM))

## Coefficient of determination
rSquared(slr)

## Basic best-fit line (without confidence bands)
ggplot(data=DFOBJ,mapping=aes(x=QVARE,y=QVARR)) +  
  geom_point(pch=21,color="black",fill="lightgray") +  
  labs(x="XXLABEL FOR QVARE##",y="XXLABEL FOR QVARE##") +  
  theme_NCStats() +  
  geom_smooth(method="lm",color="black",fill="lightgray",se=FALSE)
```

#### Inferences{-}
This assumes a linear model has been fit and assigned to `slr`.

```{r eval=FALSE, highlight=FALSE}
## Extract the coefficients table
cbind(Est=coef(slr),confint(slr))

## Extract the coefficient summary table (with default p-values)
summary(slr)

## Make a prediction with confidence/prediction intervals
nd <- data.frame(QVARE=c(NUM,NUM,...))         # data.frame to save typing
predict(slr,newdata=nd,interval="confidence")  # confidence intervals
predict(slr,newdata=nd,interval="prediction")  # prediction intervals

## Best-fit line with confidence band
ggplot(data=DFOBJ,mapping=aes(x=QVARE,y=QVARR)) +  
  geom_point(pch=21,color="black",fill="lightgray") +  
  labs(x="XXLABEL FOR QVARE##",y="XXLABEL FOR QVARE##") +  
  theme_NCStats() +  
  geom_smooth(method="lm",color="black",fill="lightgray")
```

#### Assumption Checking{-}
This assumes a linear model has been fit and assigned to `slr`.

```{r eval=FALSE, highlight=FALSE}
assumptionCheck(slr)
```

#### Transformations{-}
This assumes a linear model has been fit and assigned to `slr`.

```{r eval=FALSE, highlight=FALSE}
## Exploring different response and explanatory transformations
assumptionCheck(slr,lambday=NUM,lambdax=NUM)
```

See assumptions checking section for One-Way ANOVA above for codes to transform the variables.

&nbsp;

## Indicator Variable Regression{-}
#### Basic Indicator Variable Regression{-}
```{r eval=FALSE, highlight=FALSE}
## Fit the linear model
( ivr <- lm(QVARR~QVARE+CVAR+QVARE:CVAR,data=DFOBJ) )

## Extract coefficients table
cbind(Ests=coef(ivr),confint(ivr))

## Make a prediction
nd <- data.frame(QVARE=c(NUM,NUM),CVAR=c("VALUE","VALUE"))
predict(ivr,newdata=nd,interval="confidence")
predict(ivr,newdata=nd,interval="prediction")

## Basic best-fit line (without confidence bands)
ggplot(data=DFOBJ,mapping=aes(x=QVARE,y=QVARR,color=CVAR)) +  
  geom_point() +  
  labs(x="XXLABEL FOR QVARE##",y="XXLABEL FOR QVARE##") +  
  theme_NCStats() +  
  geom_smooth(method="lm",se=FALSE)
```

#### Inferences{-}
Assumes that you have the ultimate full model fit and assigned to `ivr`.

```{r eval=FALSE, highlight=FALSE}
anova(ivr)
```

#### Multiple Comparisons{-}
If the slopes differ among some groups (i.e., significant interaction term in the ultimate full model) then use the following.

```{r eval=FALSE, highlight=FALSE}
mcs <- emtrends(ivr,pairwise~CVAR,var="QVARE")
( mcssum <- summary(mcs,infer=TRUE) )
```

If the slopes do not differ among some groups but there is a different in intercepts among some groups (i.e., no siginficiant interaction term but a significant factor term in the ultimate full model) then first fit a model without the interaction term.

```{r eval=FALSE, highlight=FALSE}
## !!!! Only use if slopes are equal (interaction not significant) !!!!
( ivr_noint <- lm(QVARR~QVARE+CVAR,data=DFOBJ) )
```

Then perform Tukey's test on the factor term.

```{r eval=FALSE, highlight=FALSE}
mci <- emmeans(ivr_noint,pairwise~CVAR)
( mcisum <- summary(mci,infer=TRUE) )
```

&nbsp;

## Logistic Regression{-}
#### Basic Logistic Regression{-}
```{r eval=FALSE, highlight=FALSE}
## Convert categorical response to binary variable
## !!!! VALUE should be replaced with the name of the success outcome
DFOBJ$CVAR01 <- ifelse(DFOBJ$CVAR=="VALUE",1,0)

## Fit the linear model
logreg <- glm(CVAR01~QVAR,data=DFOBJ,family=binomial)

## Relationship test
anova(logreg,test="LRT")

## Perform bootstraps
b_logreg <- car::Boot(logreg)

## Create table of model coefficients
cbind(Ests=coef(logreg),confint(b_logreg,type="perc"))
```

#### Predictions{-}
Must create the following two functions first (copy these exactly).

```{r eval=FALSE, highlight=FALSE}
predProb <- function(x,alpha,beta) exp(alpha+beta*x)/(1+exp(alpha+beta*x))
predX <- function(p,alpha,beta) (log(p/(1-p))-alpha)/beta
```

This assumes that you have bootstrapped your logistic regression model and assigned the results to `b_logreg`.

```{r eval=FALSE, highlight=FALSE}
## Predict the probability of success at a given value of QVAR
p_prob <- predProb(NUM,b_logreg$t[,1],b_logreg$t[,2])
( ci_prob <- quantile(p_prob,c(0.025,0.975),type=1) )

## Predict the value of QVAR that has a certain probability of success
p_x <- predX(NUM,b_logreg$t[,1],b_logreg$t[,2])
( ci_x <- quantile(p_x,c(0.025,0.975),type=1) )
```

#### Fitted Line Plot{-}

```{r eval=FALSE, highlight=FALSE}
ggplot(data=DFOBJ,mapping=aes(x=QVAR,y=CVAR01)) +
  geom_point(alpha=0.01) +
  geom_smooth(method="glm",method.args=list(family=binomial)) +
  labs(x="XXLABEL FOR QVARXX",y="Probability of XXLABEL FOR SUCCESSXX") +
  theme_NCStats()
```

You may need to adjust the `alpha=` value higher if the points are too light.

&nbsp;
